Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-30 paper.tex
Layout tweaks

One of the key benefits of Zarr is its cloud-native design, 
but it also works well on standard file systems, where 
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
benchmarks in the following sections. 
(See \citep{durbin2020task,moore2021ome,gowan2022using} for Zarr
benchmarks in cloud settings.)
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
on compression performance, which uses a custom file format 
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data 
from subsets of a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so provides 
a reasonable and easily reproducible data point 
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic 
(see the Genomics England case-study
for benchmarks on a large human dataset that includes 
many such fields). 
Note, however, that such minimal, genotype-only data 
is something of a best-case scenario for specialised genotype
compression methods using row-based storage.

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantages at 
larger sample sizes). Zarr and Savvy have 
almost identical compression performance in this example.
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

\subsection{Subsetting the genotype matrix}
As datasets grow ever larger, the ability to efficiently access subsets 
of the data becomes increasingly important. VCF/BCF achieve efficient 
access to the data for genomic ranges 
by compressing blocks of adjacent records using \texttt{bgzip},
and storing secondary indexes alongside the original 
files with a conventional suffix~\citep{li2011tabix}. 
Thus, for a given range query we 
decompress only the necessary blocks and can quickly access
the required records. 
Methods like Savvy and XSI largely follow the same principles [TODO check and 
be more concrete]. The row-wise nature of these methods, however, mean 
that we cannot efficiently subset \emph{by sample}
(e.g., to calculate statistics within a particular cohort). In the extreme
case, if we want to access only the genotypes for a single sample
we must still retrieve and decompress the entire dataset.

\begin{figure}
\includegraphics{figures/subset-matrix-compute}
\caption{Compute performance on subsets of the matrix.
Total CPU time required to run the af-dist calculation for
a contiguous subset of 10,000 variants $\times$ 10 samples 
from the middle of the matrix
for the data in Fig~\ref{fig-data-storage}.
The \texttt{genozip} and \texttt{bcftools} pipelines involve
multiple commands required to correctly calculate the AF INFO field
required by \texttt{bcftools +af-dist}. See the Methods for full details
on the steps performed.
\label{fig-subset-matrix-compute}}
\end{figure}

@@

\begin{figure}
\includegraphics{figures/column-extract}
\caption{Time to extract the genome position and write to a text file.
Total CPU time required to extract the POS field for BCF,
sav and Zarr formats
for the data in Figure~\ref{fig-data-storage}.
For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
For sav, we used the Savvy C++ API to extract position for each variant
and output text using the \texttt{std::cout} stream. For Zarr, we read 
the variant\_position array into a NumPy array, and then wrote to
a text file using the Pandas \texttt{write\_csv} method. 
Zarr CPU time is dominated by writing the text output; we also show
the time required to populate a NumPy array with the data in Zarr,
which is less than a second. Wall-clock time (dotted line) is dominated 
in this case by file I/O. Time to output text for Savvy is not significant
for $> 1000$ samples (not shown).
\label{fig-column-extract}}
\end{figure}

\subsection{Extracting fields}
We have focused on the genotype matrix up to this point, contrasting
Zarr with existing row-wise methods.
Real-world VCFs encapsulate much more than just the genotype
matrix, and can contain large numbers of additional fields.
Fig~\ref{fig-column-extract} shows the time required to extract 
the genomic position of each variant in the simulated benchmark 
dataset, which we can use as an indicative example of a per-variant 
query. Although Savvy is many times faster than \texttt{bcftools query}
here, the row-wise storage strategy that they share means that 
the entire dataset must be read into memory and 
decompressed to extract just one field from each record. Zarr
excels at these tasks: we only read and decompress the information required.

>>>>

One of the key benefits of Zarr is its cloud-native design, 
but it also works well on standard file systems, where 
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
benchmarks in the following sections. 
(See \citep{durbin2020task,moore2021ome,gowan2022using} for Zarr
benchmarks in cloud settings.)
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
on compression performance, which uses a custom file format 
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data 
from subsets of a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so it provides 
a reasonable and easily reproducible data point 
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic 
(see the Genomics England case-study
for benchmarks on a large human dataset that includes 
many such fields). 
Note, however, that such minimal, genotype-only data 
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

Fig~\ref{fig-data-storage} shows compression performance 
on up to a million samples for chromosome 21, with 
the size of the genotype-matrix encoded as 1-bit per haploid
call included for reference.
% 2bit  1689.153258 G
% vcf   81.375831 G
% = 20.75 X, 4.8%
Gzip compressed VCF performs remarkably well, compressing 
the data to around 5\% of the 
minimal binary encoding of a biallelic genotype matrix
for 1 million samples. 
BCF provides a significant improvement in compression
performance over VCF (note the log-log scale). Genozip has 
superb compression, having far smaller file sizes that the 
other methods (although somewhat losing its advantage at 
larger sample sizes). Zarr and Savvy have 
almost identical compression performance in this example.
It is remarkable that the simple approach of compressing
two dimensional chunks of the genotype matrix 
using the Zstandard compressor~\citep{collet2021rfc} and the 
bit-shuffle filter from Blosc~\cite{alted2010modern} 
(see Methods for details) produces 
compression levels competitive with the highly specialised methods
used by Savvy.

@@

\subsection{Subsetting the genotype matrix}
\begin{figure}[t]
\includegraphics{figures/subset-matrix-compute}
\caption{Compute performance on subsets of the matrix.
Total CPU time required to run the af-dist calculation for
a contiguous subset of 10,000 variants $\times$ 10 samples 
from the middle of the matrix
for the data in Fig~\ref{fig-data-storage}.
The \texttt{genozip} and \texttt{bcftools} pipelines involve
multiple commands required to correctly calculate the AF INFO field
required by \texttt{bcftools +af-dist}. See the Methods for full details
on the steps performed.
\label{fig-subset-matrix-compute}}
\end{figure}

As datasets grow ever larger, the ability to efficiently access subsets 
of the data becomes increasingly important. VCF/BCF achieve efficient 
access to the data for genomic ranges 
by compressing blocks of adjacent records using \texttt{bgzip},
and storing secondary indexes alongside the original 
files with a conventional suffix~\citep{li2011tabix}. 
Thus, for a given range query we 
decompress only the necessary blocks and can quickly access
the required records. 
Methods like Savvy and XSI largely follow the same principles [TODO check and 
be more concrete]. The row-wise nature of these methods, however, mean 
that we cannot efficiently subset \emph{by sample}
(e.g., to calculate statistics within a particular cohort). In the extreme
case, if we want to access only the genotypes for a single sample
we must still retrieve and decompress the entire dataset.

@@

\subsection{Extracting fields}
\begin{figure}
\includegraphics{figures/column-extract}
\caption{Time to extract the genome position and write to a text file.
Total CPU time required to extract the POS field for BCF,
sav and Zarr formats
for the data in Figure~\ref{fig-data-storage}.
For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
For sav, we used the Savvy C++ API to extract position for each variant
and output text using the \texttt{std::cout} stream. For Zarr, we read 
the variant\_position array into a NumPy array, and then wrote to
a text file using the Pandas \texttt{write\_csv} method. 
Zarr CPU time is dominated by writing the text output; we also show
the time required to populate a NumPy array with the data in Zarr,
which is less than a second. Wall-clock time (dotted line) is dominated 
in this case by file I/O. Time to output text for Savvy is not significant
for $> 1000$ samples (not shown).
\label{fig-column-extract}}
\end{figure}
We have focused on the genotype matrix up to this point, contrasting
Zarr with existing row-wise methods.
Real-world VCFs encapsulate much more than just the genotype
matrix, and can contain large numbers of additional fields.
Fig~\ref{fig-column-extract} shows the time required to extract 
the genomic position of each variant in the simulated benchmark 
dataset, which we can use as an indicative example of a per-variant 
query. Although Savvy is many times faster than \texttt{bcftools query}
here, the row-wise storage strategy that they share means that 
the entire dataset must be read into memory and 
decompressed to extract just one field from each record. Zarr
excels at these tasks: we only read and decompress the information required.
